---
title: "basejump"
author: "Michael Steinbaugh"
date: "`r BiocStyle::doc_date()`"
output: BiocStyle::html_document
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{basejump}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: >
  The basejump package is an infrastructure toolkit that extends the base
  functionality of Bioconductor [@Huber2015-nw]. The package leverages the S4
  object system for object-oriented programming, and defines multiple additional
  generic functions for use in genomics research. basejump provides simple,
  user-friendly functions for the acquisition of genome annotations from
  multiple online databases, including native support for Ensembl
  [@Hubbard2002-ah] and websites supporting standard FASTA and GFF file formats.
  Consistent handling of sample metadata remains a challenge for many
  bioinformatic analyses, and basejump aims to address this by providing a suite
  of sanitization functions to help standardize these variable inputs.
  Additionally, interactive read/write operations in R can be cumbersome and
  non-trivial when working with multiple data objects; here we provide
  additional functions designed for interactive use that aim to reduce friction
  and provide consistent handling of multiple common file formats used in
  genomics research.
---

```{r setup, include=FALSE, message=FALSE}
library("basejump")
ggplot2::theme_set(theme_paperwhite())
```

# Introduction

This vignette focuses on the most common user-facing functions that fall into the following categories:

- Annotation functions
- Read/write functions
- Data functions
- Plot functions
- Sanitization functions
- Math and science functions

There are additional function families whose functionality lie outside the scope of this vignette. Consult the [basejump][] website or package documentation for more information of these types of functions, which are more technical in nature and predominantly developer-facing:

- R Markdown functions
- Atomic vector functions
- Coercion methods
- Developer functions
- Assert check functions

# The S4 object system

The [basejump][] package defines generics and methods for object-oriented programming with the S4 object system, which is used extensively by the [Bioconductor][] project. These resources describe the S4 object system in detail:

- [S4 classes and methods](https://www.bioconductor.org/help/course-materials/2017/Zurich/S4-classes-and-methods.html), Martin Morgan and Hervé Pagès, 2017.
- [A quick overview of the S4 class system](https://bioconductor.org/packages/devel/bioc/vignettes/S4Vectors/inst/doc/S4QuickOverview.pdf), Hervé Pagès, 2016.
- [The S4 object system](http://adv-r.had.co.nz/S4.html), Hadley Wickham.
- [Bioconductor for Genomic Data Science](http://kasperdanielhansen.github.io/genbioconductor), Kasper Hansen, 2016.
- [S4 System Development in Bioconductor](https://www.bioconductor.org/help/course-materials/2010/AdvancedR/S4InBioconductor.pdf), Patrick Aboyoun, 2010.

Here is a current list of S4 generics and/or methods exported by the package:

```{r}
exports <- getNamespaceExports("basejump") %>%
    sort() %>%
    .[!grepl("^\\.", .)]
isS4 <- vapply(
    X = exports,
    FUN = function(x) {
        isS4(get(x, envir = asNamespace("basejump")))
    },
    FUN.VALUE = logical(1L)
)
exports[isS4]
```

In [R][], you can check whether a function is using the S4 object system with the `base::isS4()` function. Additionally, `methods::showMethods()` and `methods::getMethod()`. In contrast, to obtain information on functions using the [S3][] object system, use `utils::methods()` instead.

# Annotation functions

Obtaining versioned genome annotations quickly and reliably from online databases such as [Ensembl][] [@Hubbard2002-ah], [GENCODE][] [@Harrow2012-hx], [RefSeq][] [@Pruitt2007-vu], and the [UCSC Genome Browser][UCSC] [@Kent2002-fo] remains overly challenging. [basejump][] aims to help address this issue by providing native gene- and transcript-level annotation support by helping users interface with [AnnotationHub][] and [ensembldb][], along with a robust set of tools to parse general feature format (GFF) and general transfer format (GTF; a.k.a. GFFv2) annotation files. GFF file parsing is handled internally by the [rtracklayer][] [@Lawrence2009-zx] package. Genome annotations are returned as `GRanges` objects, defined in the [GenomicRanges][] [@Lawrence2013-vz] package, which allows for easy access to position, chromosome, strand, and additional metadata saved in the `mcols` slot of the object. `GRanges` can be coerced to a standard `data.frame` using the `BiocGenerics::as.data.frame()` function.

## Ensembl annotations

`makeGRangesFromEnsembl()` supports multiple genome builds, and versioned releases from [Ensembl][]. By default, the function will use the latest release version available from [AnnotationHub][] and [ensembldb][].

```{r}
grch38 <- makeGRangesFromEnsembl("Homo sapiens")
summary(grch38)
glimpse(grch38)
head(names(grch38))
```

Data inside a `GRanges` object can be accessed with a number of functions defined in the [GenomeInfoDb][] package.

```{r}
GenomeInfoDb::seqnames(grch38)
GenomeInfoDb::seqinfo(grch38)
IRanges::ranges(grch38)
BiocGenerics::strand(grch38)
```

The `GRanges` class is a powerful container for genomic coordinates and metadata. For example, we can easily the gene counts per chromosome.

```{r}
grch38 %>%
    GenomeInfoDb::seqnames(.) %>%
    table() %>%
    sort(decreasing = TRUE) %>%
    head(n = 24) %>%
    .[sort(names(.))]
```

Gene metadata is contained in the `mcols` slot of the object, using a `DataFrame` internally. Here'a the current list of gene-level annotation columns returned from `makeGRangesFromEnsembl()`:

```{r}
S4Vectors::mcols(grch38) %>% colnames()
S4Vectors::mcols(grch38) %>% glimpse()
```

Transcript-level annotations are also supported, using the `format = "transcripts"` argument.

```{r}
transcripts <- makeGRangesFromEnsembl(
    organism = "Homo sapiens",
    format = "transcripts"
)
summary(transcripts)
S4Vectors::mcols(transcripts) %>% glimpse()
```

### Versioned releases

Versioned releases are supported by the "`release`" argument, which currently supports back to [Ensembl][] 87. If an older [Ensembl][] release is required, resort to using a GFF file or browse the [Bioconductor][] website to see if a suitable annotation database package is available.

```{r}
grch38_90 <- makeGRangesFromEnsembl(
    organism = "Homo sapiens",
    release = 90
)
summary(grch38_90)
GenomeInfoDb::seqinfo(grch38_90)
```

### GRCh37

The legacy [Ensembl GRCh37 genome build][GRCh37] (release 75; also known as [UCSC hg19][hg19]) is also supported, but is no longer recommended for use in new analyses. Internally, support for [GRCh37][] is provided by the [EnsDb.Hsapiens.v75][] annotation database package.

```{r}
grch37 <- makeGRangesFromEnsembl(
    organism = "Homo sapiens",
    build = "GRCh37"
)
summary(grch37)
GenomeInfoDb::seqinfo(grch37)
```

### Data frame return

For convenience, if you prefer to return annotations as a `data.frame` instead of `GRanges`, we provide the `annotable()` function. Internally, this calls the `makeGRangesFromEnsembl()` function but automatically coerces the return to a standard `data.frame`, with the identifiers as the row names. For quick access to current [Ensembl][] annotations in `tibble` format, we also recommend the [annotables][] package.

```{r, message=FALSE}
genes <- annotable("Homo sapiens")
glimpse(genes)
transcripts <- annotable("Homo sapiens")
glimpse(transcripts)
```

Note that `GRanges` objects can also be coerced to a `data.frame` using the `as.data.frame()` coercion method.

```{r}
grch38 %>% as.data.frame() %>% glimpse()
```

## Gene mapping functions

Gene-to-symbol and transcript-to-gene mappings can be easily acquired with the `gene2symbol()` and `tx2gene()` functions. Both of these functions return a `data.frame`.

```{r, message=FALSE}
gene2symbol <- makeGene2symbolFromEnsembl("Homo sapiens")
glimpse(gene2symbol)
tx2gene <- makeTx2geneFromEnsembl("Homo sapiens")
glimpse(tx2gene)
```

# R session information

```{r}
sessioninfo::session_info()
```

# References

The papers and software cited in our workflows are available as a [shared library](https://paperpile.com/shared/agxufd) on [Paperpile][].

[annotables]: https://github.com/stephenturner/annotables
[AnnotationHub]: https://bioconductor.org/packages/release/bioc/html/AnnotationHub.html
[basejump]: https://steinbaugh.com/basejump
[Bioconductor]: https://bioconductor.org
[EnsDb.Hsapiens.v75]: https://bioconductor.org/packages/release/data/annotation/html/EnsDb.Hsapiens.v75.html
[Ensembl]: https://www.ensembl.org
[ensembldb]: https://doi.org/doi:10.18129/B9.bioc.ensembldb
[FASTA]: https://www.ncbi.nlm.nih.gov/BLAST/fasta.shtml
[FlyBase]: http://flybase.org
[GenomeInfoDb]: https://bioconductor.org/packages/release/bioc/html/GenomeInfoDb.html
[GFF]: https://useast.ensembl.org/info/website/upload/gff.html
[Gencode]: https://www.gencodegenes.org
[GenomicRanges]: https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html
[ggplot2]: https://ggplot2.tidyverse.org
[GRCh37]: https://grch37.ensembl.org
[hg19]: https://genome.ucsc.edu/cgi-bin/hgGateway?db=hg19
[Paperpile]: https://paperpile.com
[R]: https://www.r-project.org
[R Markdown]: https://rmarkdown.rstudio.com
[r-wormbase]: https://steinbaugh.com/wormbase
[RefSeq]: https://www.ncbi.nlm.nih.gov/refseq
[rio]: https://cran.r-project.org/package=rio
[rtracklayer]: https://bioconductor.org/packages/release/bioc/html/rtracklayer.html
[S3]: http://adv-r.had.co.nz/S3.html
[SummarizedExperiment]: https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html
[UCSC]: http://genome.ucsc.edu
[WormBase]: https://wormbase.org
