---
title: "Install RDAVIDWebService"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: github_document
---

# macOS environment

* [Homebrew](http://brew.sh)
* [RStudio](https://www.rstudio.com)
* [Java JDK](http://www.oracle.com/technetwork/java/javase/downloads/) >= 1.8
* [OpenSSL](https://www.openssl.org/source/) >= 1.0.2

## Install OpenSSL

In Terminal.app:

## Java JDK 1.8 update

The current version of the web service only works with JDK 1.8, which must be
installed from Oracle. The Mac Java JDK is version 1.6. See
[Bioconductor thread #72188](https://support.bioconductor.org/p/72188/) and
[Stack Overflow thread #31480579](http://stackoverflow.com/questions/31480579/r-david-webservice-sudden-transport-error-301-error-moved-permanently) for more information and instructions.

Use Homebrew to install Java:

## DAVID certificates

In Terminal.app:

```{bash, eval = FALSE}
sh docs/install_openssl.sh
sh docs/RDAVIDWebService.sh
```

## Update rJava from source

See [Bioconductor thread #70090](https://support.bioconductor.org/p/70090/)
for more info.

```{r rjava, eval = FALSE}
install.packages('rJava',,'http://www.rforge.net/')
```

### Check if rJava loads

```{r rstudio check, eval = FALSE}
dyn.load("/Library/Java/JavaVirtualMachines/jdk1.8.0_102.jdk/Contents/Home/jre/lib/server/libjvm.dylib")
library("rJava")
.jinit()
.jcall("java/lang/System", "S", "getProperty", "java.runtime.version")
```


## Install RDAVIDWebService

```{r RDAVIDWebService, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("RDAVIDWebService")
browseVignettes("RDAVIDWebService")
```
